---
title: "How to cite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to cite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(priogrid)
```

## Citing Data Sources

**Always cite the original data providers when using their data.** This is not only required by most data licenses (such as CC-BY), but is also an essential practice in research.

The PRIOGRID package makes citation easy. If your dataset includes PRIOGRID variables that haven't been renamed, simply use the `pgcitations()` function to retrieve the appropriate citations for all data sources in your dataset.

```{r citation, echo = TRUE, results='markup'}
variables <- c("hilda_urban", "cru_tmp")
pgcitations(variables)
```

By maintaining original variable names and using this function, you ensure proper attribution to all data providers whose work contributes to your research.

You can also just get the biblatex entries using `pgcitations(variables, as_biblatex = T)`.
