---
title: "Citations and Bibliography"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citations and Bibliography}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(priogrid)
```

## Why Attribution Matters

PRIOGRID aggregates data from many third-party providers. Most sources are licensed under Creative Commons (CC-BY) or similar terms that **require attribution**. Beyond legal compliance, citing data providers gives credit where it is due and helps readers trace the provenance of your results.

PRIOGRID makes this easy: citation metadata is embedded in the package and linked to each variable.

## Citing Variables in Your Dataset

If your dataset uses PRIOGRID variables with their original names, `pgcitations()` retrieves all required citations in one call:

```{r citations-vector, echo=TRUE, results='markup'}
variables <- c("hilda_urban", "cru_tmp")
pgcitations(variables)
```

You can also pass a data frame directly — `pgcitations()` reads the column names:

```{r citations-dataframe, eval=FALSE}
pgcitations(pg_static)        # data.frame: uses column names
pgcitations(pg_timevarying)   # same
```

### BibLaTeX Export

For LaTeX documents, export as BibLaTeX entries:

```{r biblatex, echo=TRUE, results='markup'}
pgcitations(c("hilda_urban", "cru_tmp"), as_biblatex = TRUE)
```

Copy the output into your `.bib` file or paste it directly into an Overleaf project.

## How Citation Lookup Works

The chain from variable to citation is:

```
pgvariables$name → pgvariables$source_ids → pgsources$citation_keys → REFERENCES.bib
```

1. `pgvariables` maps variable names to source UUIDs.
2. `pgsources` maps source UUIDs to BibTeX keys (`citation_keys` column).
3. `inst/REFERENCES.bib` holds the full bibliography (~200+ entries).

You can inspect this manually:

```{r manual-lookup, echo=TRUE, results='markup'}
# Find the source IDs for a variable
pgvariables[pgvariables$name == "cru_tmp", ]

# Look up citation keys in pgsources
source_id <- pgvariables[pgvariables$name == "cru_tmp", "source_ids"]
pgsources[pgsources$id == source_id, c("source_name", "citation_keys")]
```

## Working with Bibliography Entries Directly

### Get a bibliography entry by key

`get_bibliography()` returns a `RefManageR::BibEntry` object for one or more keys:

```{r get-bib, echo=TRUE, results='markup'}
get_bibliography("harrisCRUTSVersiondataset2020")
```

Pass `as_biblatex = TRUE` for raw BibLaTeX text:

```{r get-bib-latex, echo=TRUE, results='markup'}
get_bibliography("harrisCRUTSVersiondataset2020", as_biblatex = TRUE)
```

### Extract specific fields

`get_bib_element()` pulls out a single field (author, title, journal, or year):

```{r bib-element, echo=TRUE, results='markup'}
get_bib_element("harrisCRUTSVersiondataset2020", element = "author")
get_bib_element("harrisCRUTSVersiondataset2020", element = "year")
get_bib_element("harrisCRUTSVersiondataset2020", element = "title")
```

### Parse semicolon-separated keys from pgsources

Source entries can have multiple citations separated by `"; "`. `extract_bib_elements()` handles this:

```{r extract-elements, echo=TRUE, results='markup'}
# Get all citation keys for the first source
first_source_keys <- pgsources$citation_keys[1]
first_source_keys

extract_bib_elements(first_source_keys, bib_element = "author")
```

## Citing the PRIOGRID Package Itself

`pgcitations()` always prints the package citation first. To get it directly:

```{r pkg-citation, echo=TRUE, results='markup'}
citation("priogrid")
```

## Summary

| Function | Purpose |
|---|---|
| `pgcitations(vars)` | Print all citations for a set of variables |
| `pgcitations(vars, as_biblatex = TRUE)` | Export as BibLaTeX |
| `get_bibliography(keys)` | Get BibEntry objects by key |
| `get_bibliography(keys, as_biblatex = TRUE)` | Get raw BibLaTeX text |
| `get_bib_element(key, element)` | Extract author / title / year / journal |
| `extract_bib_elements(keys_str)` | Parse semicolon-separated key strings |
