---
title: "Getting Started with PRIOGRID"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PRIOGRID}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(priogrid)
```

PRIOGRID is an R package for collecting and standardizing open spatial data into a common grid format. This tutorial covers the basics: setting up the package, downloading data, and reading it into R as tabular data — no `terra` or `sf` required.

## Initial Setup

PRIOGRID stores raw downloaded data and processed outputs in a single folder on your machine. This path persists across R sessions.

```{r setup-rawfolder, eval=FALSE}
library(priogrid)

# Set the data folder (run once; persists across sessions)
pg_set_rawfolder("/path/to/your/data/folder")
```

Once set, you can retrieve it at any time:

```{r check-rawfolder, eval=FALSE}
pg_rawfolder()
```

## Configuration

PRIOGRID has a session-scoped configuration object that controls the spatial and temporal parameters of the grid. The defaults match the official PRIOGRID release:

```{r show-config}
pg_current_config()
```

| Parameter | Default | Meaning |
|---|---|---|
| `nrow` / `ncol` | 360 / 720 | Grid dimensions (0.5° cells) |
| `crs` | `"epsg:4326"` | WGS84 geographic coordinates |
| `extent` | global | -180 to 180, -90 to 90 |
| `temporal_resolution` | `"1 year"` | Annual time steps |
| `start_date` | `1850-12-31` | First measurement date |
| `end_date` | today | Last measurement date |

You can inspect individual fields:

```{r inspect-config}
cfg <- pg_current_config()
cfg$nrow
cfg$temporal_resolution
```

The configuration is only in effect for the current R session. See the [Custom Configurations](custom-config.html) vignette for how to change it.

## Downloading the Official Release

Download the current official PRIOGRID release with a single call. This downloads a zip archive and extracts it to your raw data folder:

```{r download, eval=FALSE}
download_priogrid()
```

To see what releases are available:

```{r list-releases, eval=FALSE}
download_priogrid(list_releases = TRUE)
```

The download only happens once — subsequent calls are skipped unless you set `overwrite = TRUE`.

## Reading PRIOGRID Data

Once downloaded, load the full dataset as a `data.table`. This requires no spatial libraries.

### Static Variables

Static variables do not change over time (e.g., terrain elevation, country border distances):

```{r read-static, eval=FALSE}
pg_static <- read_pg_static()
head(pg_static)
```

Each row is a PRIOGRID cell identified by `pgid`. Columns are variable names.

### Time-Varying Variables

Time-varying variables have a `measurement_date` column in addition to `pgid`:

```{r read-timevarying, eval=FALSE}
pg_tv <- read_pg_timevarying()
head(pg_tv)
```

These tables can be joined together by `pgid`, or merged with your own data.

### Example: subsetting and merging

```{r subset-example, eval=FALSE}
library(data.table)

# Subset to a specific year
pg_2020 <- pg_tv[measurement_date == as.Date("2020-12-31")]

# Join static and time-varying on pgid
pg_merged <- merge(pg_static, pg_2020, by = "pgid")
```

## Browsing Available Variables

The `pgvariables` data frame lists all variables in PRIOGRID:

```{r pgvariables}
pgvariables
```

The `static` column indicates whether a variable varies over time. Use this to filter:

```{r filter-vars}
# Static variables
pgvariables[pgvariables$static == TRUE, "name"]

# Time-varying variables
pgvariables[pgvariables$static == FALSE, "name"]
```

## Rawfolder Structure

After downloading, your data folder has this structure:

```
{rawfolder}/
├── priogrid/
│   ├── priogrid_3_0_1_05deg_yearly.zip   # Downloaded archive
│   └── releases/
│       └── 3.0.1/
│           └── 05deg_yearly/
│               ├── pg_static.parquet        # Static data (cached)
│               ├── pg_timevarying.parquet   # Time-varying data (cached)
│               └── {varname}.rds            # Individual variable rasters
├── {source_name}/{version}/{id}/            # Raw source files
└── tmp/                                     # Temporary processing files
```

The `.parquet` files are what `read_pg_static()` and `read_pg_timevarying()` read by default. The `.rds` files contain individual variables as rasters (used by `load_pgvariable()`).

## Next Steps

- [Accessing PRIOGRID as Rasters](working-with-rasters.html) — using `terra` to work with individual variables as spatial rasters
- [Citations and Bibliography](citation.html) — how to cite data providers
- [Custom Configurations](custom-config.html) — changing resolution, extent, and time period
- [Understanding PRIOGRID Metadata](metadata.html) — exploring `pgsources`, `pgvariables`, and `pgsearch()`
