---
title: "Getting Started with PRIOGRID"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PRIOGRID}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(priogrid)
```

PRIOGRID is an R package for collecting and standardizing open spatial data into a common grid format. This tutorial covers the basics: setting up the package, downloading data, and reading it into R as tabular data — no `terra` or `sf` required.

## Initial Setup

PRIOGRID stores raw downloaded data and processed outputs in a single folder on your machine. This path persists across R sessions.

```{r setup-rawfolder, eval=FALSE}
library(priogrid)

# Set the data folder (run once; persists across sessions)
pg_set_rawfolder("/path/to/your/data/folder")
```

Once set, you can retrieve it at any time:

```{r check-rawfolder, eval=FALSE}
pg_rawfolder()
```

## Configuration

PRIOGRID has a session-scoped configuration object that controls the spatial and temporal parameters of the grid. The defaults match the official PRIOGRID release:

```{r show-config}
pg_current_config()
```

| Parameter | Default | Meaning |
|---|---|---|
| `nrow` / `ncol` | 360 / 720 | Grid dimensions (0.5° cells) |
| `crs` | `"epsg:4326"` | WGS84 geographic coordinates |
| `extent` | global | -180 to 180, -90 to 90 |
| `temporal_resolution` | `"1 year"` | Annual time steps |
| `start_date` | `1850-12-31` | First measurement date |
| `end_date` | today | Last measurement date |

You can inspect individual fields:

```{r inspect-config}
cfg <- pg_current_config()
cfg$nrow
cfg$temporal_resolution
```

The configuration is only in effect for the current R session. See the [Custom Configurations](custom-config.html) vignette for how to change it.

## Downloading the Official Release

Download the current official PRIOGRID release with a single call. This downloads a zip archive and extracts it to your raw data folder:

```{r download, eval=FALSE}
download_priogrid()
```

To see what releases are available:

```{r list-releases, eval=FALSE}
download_priogrid(list_releases = TRUE)
```

The download only happens once — subsequent calls are skipped unless you set `overwrite = TRUE`.

## Reading PRIOGRID Data

Once downloaded, load the full dataset as a `data.table`. This requires no spatial libraries.

### Static Variables

Static variables do not change over time (e.g., terrain elevation, country border distances):

```{r read-static, eval=FALSE}
pg_static <- read_pg_static()
head(pg_static)
```

Each row is a PRIOGRID cell identified by `pgid`. Columns are variable names.

### Time-Varying Variables

Time-varying variables have a `measurement_date` column in addition to `pgid`:

```{r read-timevarying, eval=FALSE}
pg_tv <- read_pg_timevarying()
head(pg_tv)
```

These tables can be joined together by `pgid`, or merged with your own data.

### Subsetting at read time

All six filter arguments default to `NULL` (no filter — full dataset). When supplied, filters are
pushed down to Arrow before any data is collected, so only the requested rows and columns touch
memory. Temporal and spatial constraints compose as AND; `pgids` and `extent` compose as union.

| Argument | Type | Effect |
|---|---|---|
| `years` | integer vector | Keep only the listed calendar years (prunes hive partitions) |
| `start_date` / `end_date` | Date | Inclusive bounds on `measurement_date` |
| `pgids` | integer vector | Keep specific PRIO-GRID cell IDs |
| `extent` | `c(xmin, xmax, ymin, ymax)` in lon/lat | Bounding box resolved to cell IDs (requires `terra`) |
| `variables` | character vector | Return only the named variable columns |

```{r read-subsets, eval=FALSE}
# A single year
pg_2020 <- read_pg_timevarying(years = 2020)

# Inclusive date range
pg_decade <- read_pg_timevarying(
  start_date = as.Date("2010-01-01"),
  end_date   = as.Date("2019-12-31")
)

# Spatial bounding box (Sub-Saharan Africa)
pg_africa <- read_pg_timevarying(
  extent = c(xmin = -20, xmax = 55, ymin = -35, ymax = 40)
)

# One variable column only
pg_tmp <- read_pg_timevarying(variables = "cru_tmp")

# Combined: two years, one region, two variables
pg_sub <- read_pg_timevarying(
  years     = c(2010, 2015),
  extent    = c(xmin = -20, xmax = 55, ymin = -35, ymax = 40),
  variables = c("cru_tmp", "ucdp_ged")
)
```

### Example: merging static and time-varying data

```{r merge-example, eval=FALSE}
library(data.table)

# Load a specific year and region with push-down, then merge with static data
pg_tv <- read_pg_timevarying(
  years  = 2020,
  extent = c(xmin = -20, xmax = 55, ymin = -35, ymax = 40)
)

pg_merged <- merge(pg_static, pg_tv, by = "pgid")
```

## Browsing Available Variables

The `pgvariables` data frame lists all variables in PRIOGRID:

```{r pgvariables}
pgvariables
```

The `static` column indicates whether a variable varies over time. Use this to filter:

```{r filter-vars}
# Static variables
pgvariables[pgvariables$static == TRUE, "name"]

# Time-varying variables
pgvariables[pgvariables$static == FALSE, "name"]
```

## Rawfolder Structure

After downloading, your data folder has this structure:

```
{rawfolder}/
├── priogrid/
│   ├── priogrid_3_0_1_05deg_yearly.zip       # Downloaded archive
│   └── releases/
│       └── 3.0.1/
│           └── 05deg_yearly/
│               ├── cog/
│               │   └── {varname}.tif          # Individual variables as Cloud-Optimized GeoTIFFs
│               ├── timevarying/
│               │   └── year=YYYY/
│               │       └── part-0.parquet      # Hive-partitioned time-varying data
│               ├── pg_static.parquet           # Static data (wide table)
│               ├── pg_static.csv.gz            # Static data (CSV bundle)
│               ├── pg_timevarying.csv.gz       # Time-varying data (CSV bundle)
│               ├── pg_config.json              # Dataset manifest (grid, layout, variables)
│               └── _checksums.csv              # Per-file MD5s
├── {source_name}/{version}/{id}/               # Raw source files
└── tmp/                                        # Temporary processing files
```

`read_pg_static()` reads `pg_static.parquet`; `read_pg_timevarying()` reads the hive-partitioned `timevarying/` dataset, pushing any filters down to Arrow first. The `cog/*.tif` files hold individual variables as Cloud-Optimized GeoTIFFs and are read by `load_pgvariable()`. The `*.csv.gz` bundles provide the same tables for non-R users, and `pg_config.json` records the grid and layout.

## Next Steps

- [Accessing PRIOGRID as Rasters](working-with-rasters.html) — using `terra` to work with individual variables as spatial rasters
- [Citations and Bibliography](citation.html) — how to cite data providers
- [Custom Configurations](custom-config.html) — changing resolution, extent, and time period
- [Understanding PRIOGRID Metadata](metadata.html) — exploring `pgsources`, `pgvariables`, and `pgsearch()`
