---
title: "Understanding PRIOGRID Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding PRIOGRID Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(priogrid)
```

PRIOGRID embeds rich metadata about every variable and data source. This tutorial walks through the three core metadata objects — `pgvariables`, `pgsources`, and `pgchecksum` — and the search and discovery tools built on top of them.

## Variables: `pgvariables`

`pgvariables` is a data frame listing every variable PRIOGRID can calculate:

```{r pgvariables}
pgvariables$name
```

| Column | Description |
|---|---|
| `name` | Variable name used in `load_pgvariable()`, `calc_pg()`, etc. |
| `static` | `TRUE` = no temporal dimension; `FALSE` = time-varying |
| `source_ids` | Comma-separated UUIDs of the data sources feeding this variable |

Static and time-varying variables:

```{r split-vars}
pgvariables[pgvariables$static == TRUE,  "name"]  # terrain, borders, etc.
pgvariables[pgvariables$static == FALSE, "name"]  # climate, conflict, etc.
```

## Data Sources: `pgsources`

`pgsources` describes every raw data source that PRIOGRID draws on:

```{r pgsources}
str(pgsources)
```

Key columns:

| Column | Description |
|---|---|
| `id` | UUID — links to `pgvariables$source_ids` |
| `source_name` | Human-readable name (e.g., `"CRU TS"`, `"UCDP GED"`) |
| `source_version` | Data version used |
| `license` | Data license (e.g., `"CC BY 4.0"`) |
| `citation_keys` | Semicolon-separated BibTeX keys → `inst/REFERENCES.bib` |
| `tags` | Comma-separated tags (e.g., `"climate"`, `"conflict"`) |
| `spatial_extent` | `"World"`, `"Multiple continents"`, etc. |
| `temporal_resolution` | `"Yearly"`, `"Monthly"`, `"Static"`, etc. |
| `download_url` | Primary download URL |
| `website_url` | Landing page URL |
| `prio_mirror` | PRIO-hosted mirror URL |

Browse sources by tag or license:

```{r filter-sources}
# All sources with CC-BY license
pgsources[grepl("CC BY", pgsources$license), c("source_name", "license")]

# Yearly time-varying sources
pgsources[pgsources$temporal_resolution == "Yearly", c("source_name", "source_version")]
```

## File Integrity: `pgchecksum`

`pgchecksum` stores MD5 checksums for downloaded files, allowing you to verify that your local copies match those used to build the official release:

```{r pgchecksum}
pgchecksum
```

Use `check_pgsourcefiles()` to run the check against your local storage:

```{r check-files, eval=FALSE}
check_pgsourcefiles()
# "All files in your local storage are similar to a tested set."
```

## Metadata Linking: Variables → Sources → Bibliography

The three metadata objects form a chain:

```
pgvariables$source_ids → pgsources$id → pgsources$citation_keys → REFERENCES.bib
```

Here's the full lookup for a single variable:

```{r metadata-chain}
# 1. Find source IDs for a variable
var_row <- pgvariables[pgvariables$name == "cru_tmp", ]
source_ids <- strsplit(var_row$source_ids, ", ")[[1]]
source_ids

# 2. Look up source metadata
src <- pgsources[pgsources$id %in% source_ids, c("source_name", "source_version", "citation_keys", "license")]
src

# 3. Get citation keys
bibkeys <- unlist(strsplit(src$citation_keys, "; "))
bibkeys
```

Then retrieve the full bibliography:

```{r get-bib, echo=TRUE, results='markup'}
get_bibliography(bibkeys[1])
```

## Searching Metadata: `pgsearch()`

`pgsearch()` searches across all text fields in `pgsources` using regex:

```{r pgsearch}
results <- pgsearch("population")
```

It returns a named list with matches from each search field:

```{r pgsearch-results}
# Sources matching by name
results$in_name[, c("source_name", "source_version")]

# Sources matching by tag
results$in_tags[, c("source_name", "tags")]
```

Search by temporal resolution:

```{r pgsearch-temporal}
pgsearch("Monthly")$in_temporal_resolution[, c("source_name", "temporal_resolution")]
```

Search by spatial extent:

```{r pgsearch-spatial}
pgsearch("World")$in_spatial_extent[, c("source_name", "spatial_extent")]
```

### Searching Bibliography Elements

Include `bib_element` to also search author names, titles, journals, or years:

```{r pgsearch-bib, eval=TRUE, results='markup'}
# Find sources citing a specific author
harris_results <- pgsearch("Harris", bib_element = "author")
harris_results$in_element[, c("source_name", "citation_keys")]
```

## Listing Raw Files

`pg_rawfiles()` returns a data frame of all files PRIOGRID can download, with their URLs and expected local paths:

```{r rawfiles}
files <- pg_rawfiles()
head(files[, c("source_name", "source_version", "filename")])
```

Filter to a specific source:

```{r rawfiles-filter}
files[files$source_name == "ETH ICR cShapes", c("source_name", "filename", "url")]
```

Download a specific source:

```{r download-source, eval=FALSE}
ucdp_files <- pg_rawfiles() |> dplyr::filter(source_name == "UCDP GED")
download_pg_rawdata(file_info = ucdp_files)
```

Get the local path to a downloaded file:

```{r get-pgfile, eval=FALSE}
get_pgfile(
  source_name    = "ETH ICR cShapes",
  source_version = "2.0",
  id             = "ec3eea2e-6bec-40d5-a09c-e9c6ff2f8b6b"
)
```

## Checking Data Availability

`pg_data_availability()` shows which sources have been downloaded locally:

```{r data-availability, eval=FALSE}
pg_data_availability()
#   source_name   source_version n_files n_present all_present
# 1 CRU TS        4.08                1         1        TRUE
# 2 ETH ICR cShapes 2.0              1         1        TRUE
# ...
```

## Next Steps

- [Citations and Bibliography](citation.html) — citing data providers in publications
- [Contributing to PRIOGRID](contributing.html) — adding new sources and variables
